Left-linear rules in adhesive categories

### 2024-02-06 icalp/fusione.tex
cose

%\todo{A VERY NICE INTRODUCTION}
%
%SOME IDEAS (BADLY WRITTEN, JUST TO START TO PUT IDEAS IN ORDER)

One of the key pay-off of concurrency theory is the idea that the behaviour of 
a computational device can be modelled by abstracting its sequences of steps
%performed by the device 
via a suitable equivalence. Intuitively, this equivalence 
captures when steps are causally unrelated and thus could be executed in any 
order, and possibly concurrently. 
%
Two of the seminal contribution to this line of research have been
Mazurkiewicz's traces~\cite{xxx} and Winskel's event structures~\cite{xxx}
%
Concerning formalisms based on the rewriting paradigm, i.e. where the
behaviour is dictated by a set of rules which have to be instantiated with
the system at hand, concurrency often boils down to having a notion of 
independence between two consecutive steps.
%thus viewing a sequence of steps as a concurrent computation. 
As noteworthy examples, we have 
interchange law~\cite{Mes92} in term rewriting and permutation 
equivalence~\cite{JJL80} in $\lambda$-calculi.

Among rewriting-based formalisms, those manipulating graph-like structures
proved useful in many settings for representing the dynamics of systems: the 
graph serves to represent the entities and their relations within states, while 
rewriting steps, modifying the graph, model computation steps that results 
in changes to the system state. The DPO (double-pushout) approach~\cite{xxx} is
nowadays considered the standard one for these structures, mostly thanks to its 
flexibility: the rules allow to specify that, 
in a rewriting step, some entities in the current state are removed, some others are 
needed for the rewriting step to occur but remain unaltered and some new entities 
can be generated. Concerning concurrency, the chosen notion is shift equivalence,
and independence is formally captured by the so-called Church-Rosser 
theorem of DPO rewriting~\cite{CMREHL:AAGT}.

A noteworthy advantage is that DPO rewriting can be formulated over
adhesive categories and their variants~\cite{lack2005adhesive,ehrig2006weak}. 
%
Operating within the framework of adhesive categories allows one to devise 
the foundations of a rewriting theory which can be subsequently instantiated 
to the context of interest, thus avoiding the need of repeatedly proving
similar ad-hoc results in for each specific setting. In fact, the framework is
so general that it allowed to recast in the DPO approach many additional formalisms,
ranging from xxx to string diagrams~\cite{bonchi2022string}.

So far, most of the theoretical results focussed on what are called \emph{linear rules}: 
intuitively, a step is required only to consume, preserve and generate entities. 
However, in some situations it is also necessary to ``merge'' parts of the state. 
This happens already in the context of string diagram mentioned above, but also 
in graphical implementations of nominal calculi where, as a result of name passing, 
the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the visual modelling of bonding in
biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms is
at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
explicitly merging nodes corresponding to equal expression, the idea
here - conceptually and, as we will see, also mathematically similar -
is to maintain an equivalence\todo{P: more precisely a congruence}
over the nodes.  \todo{P: la sequenza di lavori su questo \`e ampia,
  vedere se c'\`e qualcosa d'altro}

=========DA QUI SUPERSEDED\\
Rewriting systems over graphs are used in uncountably many setting for
representing the dynamics of software systems: the graph serves to represent
the entities and their relations within states , while rewriting steps,
modifying the graph, model computational steps that results in
changes to the system state.

@@

In some situations rewriting with linear rules are not sufficiently
expressive for capturing the dynamic of a system. This is the case
when a computational step is required not only to consume, preserve
and generated, but also ``merge'' parts of the state. This happens,
e.g., in nominal calculi where, as a result of name passing, the
received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the modelling of bonding in
biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms is
at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}. Rather than
explicitly merging nodes corresponding to equal expression, the idea
here - conceptually and, as we will see, also mathematically similar -
is to maintain an equivalence\todo{P: more precisely a congruence}
over the nodes.  \todo{P: la sequenza di lavori su questo \`e ampia,
  vedere se c'\`e qualcosa d'altro}\\
=========

@@

\begin{definition}
	Let $\X$ be a category and $\mathcal{A}$ a  subclass of $\mor(\X)$. We say that  $\mathcal{A}$ is:
	\begin{itemize}
		\parbox{11cm}{\item 
		\emph{stable under pushouts (pullbacks)} if for every pushout (pullbacks) square as the one on the right, 	if $m \in \mathcal{A}$ ($n\in \mathcal{A}$) then $n \in \mathcal{A}$ ($m \in \mathcal{A}$);
			\item \emph{closed under composition} if $g, f\in \mathcal{A}$ implies $g\circ f\in \mathcal{A}$ whenever $g$ and $f$ are composable; }\hfill
		\parbox{2cm}{$\xymatrix{A\ar[r]^f  \ar[d]_{m}& B \ar[d]^n \\ C \ar[r]_g & D}$}
		\parbox{11cm}{}\hfill

		\item \emph{closed under decomposition} if $g\circ f, g\in \mathcal{A}$ implies $f\in \mathcal{A}$.
	\end{itemize}
\end{definition}

\rem{rimuoverei}{
\begin{remark}
  In the previous definition, ``decomposition'' corresponds to
  ``left cancellation'', but we prefer to stick to the name commonly
  used in literature (see e.g.~\cite{habel2012mathcal}).
\end{remark}
}

@@

\rem{Qui l'idea \`e mostrare che valgono per $\mathcal{M}$ le naturali generalizzazioni delle propriet\`a delle adesive. Credo si possa dirle in prosa e metterle in appendice.}{%
A first result we can prove regards closure under decomposition of $\mathcal{M}$.

\begin{corollary}\label{prop:bal}
If $\X$ is an $\mathcal{M}$-adhesive categories, then every epimorphisms in $\mathcal{M}$ is an isomorphisms. In particular, every adhesive category $\X$ is \emph{balanced}: if a morphism is monic and epic, then it is an isomorphism.
\end{corollary}
}

\todo{P: qui probabilmente si puo' compattare tenendo solo cio' che si usa}
$\mathcal{M}$-adhesivity is well-behaved with respect to  the comma construction \cite{mac2013categories}, as shown by the following theorem.
\begin{theorem}[\cite{ehrig2006fundamentals,lack2005adhesive}]\label{lem:comma}
	Let $\A$ and $\B$ be respectively an $\mathcal{M}$-adhesive and an $\mathcal{M}'$-adhesive category. Let also $L:\A\rightarrow \C$ be a functor that preserves $\mathcal{M}$-pushouts, and  $R:\B\rightarrow \C$ be a functor which preserves pullbacks.Then $\comma{L}{R}$ is $\cma{M}{M'}$-adhesive, where 
	\[
	\cma{M}{M}':=\{(h,k)\in \mathcal{A}(\comma{L}{R}) \mid h\in \mathcal{M}, k\in \mathcal{M}'\}\]
\end{theorem}

@@

\rem{Questo penso tutto in appendice, eventualmente tenendo la decomposition o richiamandola quando si parla di switch}{%
We end this section recalling two properties of $\mathcal{M}$-adhesive categories:  $\mathcal{M}$-pushout-pullback decomposition and uniqueness of pushouts complements.

\begin{theoremEnd}[category=sec2]{lemma}[$\mathcal{M}$-pushout-pullback decomposition]\label{lem:popb} Let $\X$ be an $\mathcal{M}$-adhesive category  and suppose that, in the diagram below, the whole rectangle is a pushout and the right square a pullback.
\[\xymatrix{X \ar[d]_{a} \ar[r]^{f}& \ar[r]^{g} Y \ar[d]^{b}& Z \ar[d]^{c}\\ A \ar[r]_{h}& B \ar[r]_{k}& C}\]
	Then the following statements hold true:
	\begin{enumerate}
\item if $a$ belongs to $\mathcal{M}$ and $k$ is a monomorphism,  then both squares are pushouts and pullbacks;
\item if $f$ and $k $ are in  $\mathcal{M}$, then both squares are pushouts and pullbacks.
	\end{enumerate}
\end{theoremEnd}
\begin{proof}\begin{proofEnd}
		\begin{enumerate}
			\item By \Cref{prop:stab}, it follows that both squares are pushouts, thus the thesis follows from \Cref{prop:pbpoad}.
			\item By hypothesis, $g$ is the pullback of an arrow in $\mathcal{M}$, thus it belongs to it. But then $g\circ f\in \mathcal{M}$ too  and the whole rectangle is a $\mathcal{M}$-pushout. Therefore, by \Cref{prop:pbpoad} a pullback, so that its left half is a pullback too, by \Cref{prop:pbpo}. Moreover $k\circ h$ is in $\mathcal{M}$ as the pushout of $g\circ f$ and, by \Cref{cor:deco}, we also know that $h\in \mathcal{M}$.

			Using \Cref{lem:po1}, it is enough to show that the left half of the original rectangle is a pushout. We can build the following cube:
			\[\xymatrix@C=20pt@R=10pt{ & &X\ar@{>->}[dl]_{f} \ar[ddd]^(.333333){\id{X}}|(.666666)\hole\ar[rrr]^{a} &&&A \ar[ddd]^{\id{A}} \ar@{>->}[dl]^{h}\\& Y\ar[dl]_{\id{Y}} \ar[ddd]|(.333333)\hole_{\id{Y}} &&&B \ar[ddd]^{\id{B}} \ar[dl]_{\id{B}} \\Y \ar@{>->}[ddd]_{g} \ar[rrr]^{b}&&& B\ar@{>->}[ddd]^{k}\\&&X \ar[rrr]|(.34)\hole^{a}|(.67)\hole \ar@{>->}[dl]_{f}&&& A \ar@{>->}[dl]^{h}\\ & Y \ar[rrr]|(.67)\hole^{b} \ar@{>->}[dl]_{g}&&& B \ar@{>->}[dl]^{k}\\ Z\ar[rrr]_{c} &&& C}\]
			Its vertical faces are all pullbacks and all the vertical arrows are in $\mathcal{M}$, hence the top face is a pushout and we can conclude. \qedhere 
		\end{enumerate}
	\end{proofEnd}
\end{proof}

	The claim now follows since $k_2$ is a monomorphism.
	\end{proofEnd} 
\end{proof}
}
\subsection{DPO rewriting systems derivations}
$\mathcal{M}$-adhesive categories are the right context in which to perform abstract rewriting using the so-called ``douple pushout approach'' (DPO). We will recall the basic definitions and properties of this approach to abstract rewriting.

@@

@@

\appendix

\section{A large class of tame rewriting systems}\label{app:fill}
 In \Cref{sec:ade} we introduced the notion of tame left-linear DPO rewriting system, a class that contains, for instance, all the linear rewriting systems. This result can be further refined: in a\cite{baldan2011adhesivity} a class  $\mathbb{B}$ of (quasi)adhesive category is defined for which the local Church-Rosser Theorem holds for left-linear DPO rewriting system. In our language, this means that every left-linear DPO rewriting system based on a category in $\mathbb{B}$ is tame. This section is devoted to repropose, and slightly generalize, the results of \cite{baldan2011adhesivity} to our context.

@@

\subsection{Some results on $\mathcal{A}$-stable and $\mathcal{A}$-Van Kampen squares}
We will start proving some general results regarding $\mathcal{A}$-Van Kampen and $\mathcal{A}$-stable squares, which, wll be used in the proof of the results of \Cref{sec:ade}. Let us begin recalling some classical results about  about pullbacks and pushouts

>>>>

\todo{A VERY NICE INTRODUCTION}

SOME IDEAS (BADLY WRITTEN, JUST TO START TO PUT IDEAS IN ORDER)

Rewriting systems over graphs are used in uncountably many setting for
representing the dynamics of software systems: the graph serves to represent
the entities and their relations within states , while rewriting steps,
modifying the graph, model computational steps that results in
changes to the system state.

@@

In some situations rewriting with linear rules are not sufficiently
expressive for capturing the dynamic of a system. This is the case
when a computational step is required not only to consume, preserve
and generated, but also ``merge'' parts of the state. This happens,
e.g., in nominal calculi where, as a result of name passing, the
received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the modelling of bonding in
biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms is
at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimizations~\cite{WNW:egg}. Rather than
explicitly merging nodes corresponding to equal expression, the idea
here - conceptually and, as we will see, also mathematically similar -
is to maintain an equivalence\todo{P: more precisely a congruence}
over the nodes.  \todo{P: la sequenza di lavori su questo \`e ampia,
  vedere se c'\`e qualcosa d'altro}

\todo{P: something about string diagrams?}

@@

\begin{definition}
	Let $\X$ be a category and $\mathcal{A}$ a  subclass of $\mor(\X)$. We say that  $\mathcal{A}$ is:
	\begin{itemize}
		\parbox{11cm}{\item 
		\emph{stable under pushouts (pullbacks)} if for every pushout (pullbacks) square as the one on the right, 	if $m \in \mathcal{A}$ ($n\in \mathcal{A}$) then $n \in \mathcal{A}$ ($m \in \mathcal{A}$);
			\item \emph{closed under composition} if $g, f\in \mathcal{A}$ implies $g\circ f\in \mathcal{A}$ whenever $g$ and $f$ are composable.}\hfill
		\parbox{2cm}{$\xymatrix{A\ar[r]^f  \ar[d]_{m}& B \ar[d]^n \\ C \ar[r]_g & D}$}
		\parbox{11cm}{}\hfill 
	\end{itemize}
\end{definition}

@@

\todo{P: qui probabilmente si puo' compattare tenendo solo cio' che si usa}
\todo{Si usa tutto per la classe B+ [DC]}
$\mathcal{M}$-adhesivity is well-behaved with respect to  the comma construction \cite{mac2013categories}, as shown by the following theorem.
\begin{theorem}[\cite{ehrig2006fundamentals,lack2005adhesive}]\label{lem:comma}
	Let $\A$ and $\B$ be respectively an $\mathcal{M}$-adhesive and an $\mathcal{M}'$-adhesive category. Let also $L:\A\rightarrow \C$ be a functor that preserves $\mathcal{M}$-pushouts, and  $R:\B\rightarrow \C$ be a functor which preserves pullbacks.Then $\comma{L}{R}$ is $\cma{M}{M'}$-adhesive, where 
	\[
	\cma{M}{M}':=\{(h,k)\in \mathcal{A}(\comma{L}{R}) \mid h\in \mathcal{M}, k\in \mathcal{M}'\}\]
\end{theorem}

@@

We will prove in \Cref{app:ade} two key properties of $\mathcal{M}$-adhesive categories,  $\mathcal{M}$-pushout-pullback decomposition and uniqueness of pushouts complements, which will be used in this paper.

\subsection{DPO rewriting systems derivations}
$\mathcal{M}$-adhesive categories are the right context in which to perform abstract rewriting using the so-called ``douple pushout approach'' (DPO). We will recall the basic definitions and properties of this approach to abstract rewriting.

@@

\todo[inline]{Io invece la introdurrei proprio come punto teorico che mi sembra parecchio importante: qua li condizioni mi garantiscono che gli independence pair implicano la scambiabilità? [DC]}

@@

\appendix
\section{Properties of $\mathcal{M}$-adhesive categories}\label{app:ade}
\subsection{Some results on $\mathcal{A}$-stable and $\mathcal{A}$-Van Kampen squares}
We will start proving some general results regarding $\mathcal{A}$-Van Kampen and $\mathcal{A}$-stable squares, which, wll be used in the proof of the results of \Cref{sec:ade}. Let us begin recalling some classical results about  about pullbacks and pushouts

\subsection{Useful properties of $\mathcal{M}$-adhesive categories}

We can begin with an observation about decompo

\begin{corollary}\label{prop:bal}
	If $\X$ is an $\mathcal{M}$-adhesive categories, then every epimorphisms in $\mathcal{M}$ is an isomorphisms. In particular, every adhesive category $\X$ is \emph{balanced}: if a morphism is monic and epic, then it is an isomorphism.
\end{corollary}

\begin{lemma}[$\mathcal{M}$-pushout-pullback decomposition]\label{lem:popb} Let $\X$ be an $\mathcal{M}$-adhesive category  and suppose that, in the diagram below, the whole rectangle is a pushout and the right square a pullback.
	\[\xymatrix{X \ar[d]_{a} \ar[r]^{f}& \ar[r]^{g} Y \ar[d]^{b}& Z \ar[d]^{c}\\ A \ar[r]_{h}& B \ar[r]_{k}& C}\]
	Then the following statements hold true:
	\begin{enumerate}
		\item if $a$ belongs to $\mathcal{M}$ and $k$ is a monomorphism,  then both squares are pushouts and pullbacks;
		\item if $f$ and $k $ are in  $\mathcal{M}$, then both squares are pushouts and pullbacks.
	\end{enumerate}
\end{lemma}
\begin{proof}		\begin{enumerate}
			\item By \Cref{prop:stab}, it follows that both squares are pushouts, thus the thesis follows from \Cref{prop:pbpoad}.
			\item By hypothesis, $g$ is the pullback of an arrow in $\mathcal{M}$, thus it belongs to it. But then $g\circ f\in \mathcal{M}$ too  and the whole rectangle is a $\mathcal{M}$-pushout. Therefore, by \Cref{prop:pbpoad} a pullback, so that its left half is a pullback too, by \Cref{prop:pbpo}. Moreover $k\circ h$ is in $\mathcal{M}$ as the pushout of $g\circ f$ and, by \Cref{cor:deco}, we also know that $h\in \mathcal{M}$.

			Using \Cref{lem:po1}, it is enough to show that the left half of the original rectangle is a pushout. We can build the following cube:
			\[\xymatrix@C=20pt@R=10pt{ & &X\ar@{>->}[dl]_{f} \ar[ddd]^(.333333){\id{X}}|(.666666)\hole\ar[rrr]^{a} &&&A \ar[ddd]^{\id{A}} \ar@{>->}[dl]^{h}\\& Y\ar[dl]_{\id{Y}} \ar[ddd]|(.333333)\hole_{\id{Y}} &&&B \ar[ddd]^{\id{B}} \ar[dl]_{\id{B}} \\Y \ar@{>->}[ddd]_{g} \ar[rrr]^{b}&&& B\ar@{>->}[ddd]^{k}\\&&X \ar[rrr]|(.34)\hole^{a}|(.67)\hole \ar@{>->}[dl]_{f}&&& A \ar@{>->}[dl]^{h}\\ & Y \ar[rrr]|(.67)\hole^{b} \ar@{>->}[dl]_{g}&&& B \ar@{>->}[dl]^{k}\\ Z\ar[rrr]_{c} &&& C}\]
			Its vertical faces are all pullbacks and all the vertical arrows are in $\mathcal{M}$, hence the top face is a pushout and we can conclude. \qedhere 
		\end{enumerate}
\end{proof}

		The claim now follows since $k_2$ is a monomorphism.
	\end{proofEnd} 
\end{proof}

\section{$\mathcal{M}$-adhesivity is not enough}\label{app:fill}
 In \Cref{sec:ade} we introduced the notion of tame left-linear DPO rewriting system, a class that contains, for instance, all the linear rewriting systems. This result can be further refined: in a\cite{baldan2011adhesivity} a class  $\mathbb{B}$ of (quasi)adhesive category is defined for which the local Church-Rosser Theorem holds for left-linear DPO rewriting system. In our language, this means that every left-linear DPO rewriting system based on a category in $\mathbb{B}$ is tame. This section is devoted to repropose, and slightly generalize, the results of \cite{baldan2011adhesivity} to our context.

@@
